Text embedding models yield high-resolution insights into conceptual knowledge from short multiple-choice quizzes

### 2023-03-01 paper/main.tex
JRM final pass through

\begin{abstract}\noindent We develop a mathematical framework, based on natural
language processing models, for tracking and characterizing the acquisition of
conceptual knowledge. Our approach embeds each concept in a high-dimensional
representation space, where nearby coordinates reflect similar or related
concepts. We test our approach using behavioral data from participants who
answered small sets of multiple-choice quiz questions interleaved between
watching two course videos from the Khan Academy platform. We apply our
framework to the videos' transcripts and the text of the quiz questions to
quantify the content of each moment of video and each quiz question. We use
these embeddings, along with participants' quiz responses, to track how the
learners' knowledge changed after watching each video. Our findings show how a
small set of quiz questions may be used to obtain rich and meaningful, high-resolution
insights into what each learner knows, and how their knowledge
changes over time as they learn.

@@

We also wrote a set of multiple-choice quiz questions that we hoped would
enable us to evaluate participants' knowledge about each individual lecture,
along with related knowledge about physics concepts not specifically presented in either
video (see Supplementary Tab.~\questions~for the full list of questions in our stimulus
pool). Participants answered questions randomly drawn from each content area
(Lecture~1, Lecture~2, and general physics knowledge) on each of the three
quizzes. Quiz~1 was intended to assess participants' ``baseline'' knowledge
before training, Quiz~2 assessed knowledge after watching the \textit{Four
Fundamental Forces} video (i.e., Lecture~1), and Quiz~3 assessed knowledge
after watching the \textit{Birth of Stars} video (i.e., Lecture~2).

@@

    \caption{\textbf{Modeling course content.} \textbf{A. Building a document pool
    from sliding windows of text.} We decompose each lecture's transcript into a
    series of overlapping sliding windows. The full set of transcript snippets (across
    all windows) may be treated as a set of ``documents'' for training a text
    embedding model. \textbf{B. Constructing lecture content
    \textit{trajectories}.} After training the model on the sliding windows from
    both lectures, we transform each lecture into a ``trajectory'' through text
    embedding space by joining the embedding coordinates of successive sliding
    windows parsed from its transcript. \textbf{C. Embedding multiple lectures and
    questions in a shared space.} We apply the same model (trained on the two
    lectures' windows) to both lectures, along with the text of each question in
    our pool (Supplementary Tab.~\questions), to project them into a shared text embedding space.
    This results in one trajectory per lecture and one coordinate for each
    question. Here, we have projected the 15-dimensional embeddings onto their first
    3 principal components for visualization.}

@@

We hypothesized that a topic model trained on transcripts of the two lectures
should also capture the conceptual knowledge probed by each quiz question. If
indeed the topic model could capture information about the deeper conceptual
content of the lectures (i.e., beyond surface-level details such as particular
word choices), then we should be able to recover a correspondence between each
lecture and questions \textit{about} each lecture. Importantly, such a
correspondence could not solely arise from superficial text matching between
lecture transcripts and questions, since the lectures and questions used
different words. Simply comparing the average topic weights from each lecture
and question set (averaging across time and questions, respectively) reveals a
striking correspondence (Supplementary Fig.~\topicWeights). Specifically, the average topic
weights from Lecture~1 are strongly correlated with the average topic weights
from Lecture~1 questions ($r(13) = 0.809,~p < 0.001$, 95\% confidence interval
(CI)~$= [0.633,~0.962]$), and the average topic weights from Lecture~2 are
strongly correlated with the average topic weights from Lecture~2 questions
($r(13) = 0.728,~p = 0.002$, 95\% CI~$= [0.456,~0.920]$). At the same time, the
average topic weights from the two lectures are \textit{negatively} correlated with
their non-matching question sets (Lecture~1 video vs.~Lecture~2 questions: $r(13) =
-0.547,~p = 0.035$, 95\% CI~$= [-0.812, ~-0.231]$; Lecture~2 video vs.~Lecture~1 questions: $r(13)
= -0.612,~p = 0.015$, 95\% CI~$= [-0.874,~-0.281]$), indicating that the topic model
also exhibits some degree of specificity. The full set of pairwise comparisons
between average topic weights for the lectures and question sets is reported in
Supplementary Figure~\topicWeights.

@@

If we are able to accurately estimate a participant's knowledge about the
content tested by a given question, our estimates of their knowledge should carry some
predictive information about whether the participant is likely to answer that
question correctly or incorrectly. We developed a statistical approach to test this claim.
For each question in turn, we used equation~\ref{eqn:prop} to estimate each
participant's knowledge at the given question's embedding space coordinate,
using all \textit{other} questions that participant answered on the same quiz.
%For each question in turn, for each
%participant, we used Equation~\ref{eqn:prop} to estimate (using all
%\textit{other} questions from the same quiz, from the same participant) the
%participant's knowledge at the held-out question's embedding coordinate.
For
each quiz, we grouped these estimates into two distributions: one for the
estimated knowledge at the coordinates of \textit{correctly} answered
questions, and another for the estimated knowledge at the coordinates of
\textit{incorrectly} answered questions (Fig.~\ref{fig:predictions}). We then
used independent samples $t$-tests to compare the means of these distributions
of estimated knowledge.

@@

We transformed each sliding window's text into a topic vector, and then used
linear interpolation (independently for each topic dimension) to resample the
resulting timeseries to one vector per second. We also used the fitted model to
obtain topic vectors for each question in our pool (see Supplementary Tab.~\questions). Taken
together, we obtained a \textit{trajectory} for each video, describing its path
through topic space, and a single coordinate for each question
(Fig.~\ref{fig:sliding-windows}C). Embedding both videos and all of the
questions using a common model enables us to compare the content from different
moments of videos, compare the content across videos, and estimate potential
associations between specific questions and specific moments of video.

@@

Intuitively, $\mathrm{ncorr}(x, y)$ is the correlation between two topic
vectors (e.g., the topic vector from one timepoint in a lecture, $x$, and the
topic vector for one question, $y$), normalized by the minimum and maximum
correlations (across all timepoints $t$ and questions $Q$) to range between 0 and 1,
inclusive. Equation~\ref{eqn:prop} then computes the weighted average
proportion of correctly answered questions about the content presented at
timepoint $t$, where the weights are given by the normalized correlations
between timepoint $t$'s topic vector and the topic vectors for each question.
The normalization step (i.e., using $\mathrm{ncorr}$ instead of the raw
correlations) insures that every question
contributes some non-zero amount to the knowledge estimate.

@@

An important feature of our approach is that, given a trained text embedding
model and participants' quiz performance on each question, we can estimate
their knowledge about \textit{any} content expressible by the embedding
model---not solely the content explicitly probed by the quiz questions, or even
appearing in the lectures. To visualize these estimates
(Fig.~\ref{fig:knowledge-maps}, Supplementary Figs.~\individualKnowledgeMapsA,~\individualKnowledgeMapsB,~\individualKnowledgeMapsC,~\individualLearningMapsA,
and~\individualLearningMapsB), we used Uniform Manifold Approximation and
Projection (UMAP; \citealp{McInEtal18a, McInEtal18b}) to construct a 2D projection of the
text embedding space. Sampling the original 100-dimensional space at high
resolution to obtain an adequate set of topic vectors spanning the embedding
space would be computationally intractable. However, sampling a 2D grid is
trivial.

>>>>

\begin{abstract}\noindent We develop a mathematical framework, based on natural
language processing models, for tracking and characterizing the acquisition of
conceptual knowledge. Our approach embeds each concept in a high-dimensional
representation space, where nearby coordinates reflect similar or related
concepts. We test our approach using behavioral data from participants who
answered small sets of multiple-choice quiz questions interleaved between
watching two course videos from the Khan Academy platform. We apply our
framework to the videos' transcripts and the text of the quiz questions to
quantify the content of each moment of video and each quiz question. We use
these embeddings, along with participants' quiz responses, to track how the
learners' knowledge changed after watching each video. Our findings show how a
small set of quiz questions may be used to obtain rich and meaningful
high-resolution insights into what each learner knows, and how their knowledge
changes over time as they learn.

@@

We also wrote a set of multiple-choice quiz questions that we hoped would
enable us to evaluate participants' knowledge about each individual lecture,
along with related knowledge about physics concepts not specifically presented in either
video (see Supp. Tab.~\questions~for the full list of questions in our stimulus
pool). Participants answered questions randomly drawn from each content area
(Lecture~1, Lecture~2, and general physics knowledge) on each of the three
quizzes. Quiz~1 was intended to assess participants' ``baseline'' knowledge
before training, Quiz~2 assessed knowledge after watching the \textit{Four
Fundamental Forces} video (i.e., Lecture~1), and Quiz~3 assessed knowledge
after watching the \textit{Birth of Stars} video (i.e., Lecture~2).

@@

    \caption{\textbf{Modeling course content.} \textbf{A. Building a document pool
    from sliding windows of text.} We decompose each lecture's transcript into a
    series of overlapping sliding windows. The full set of transcript snippets (across
    all windows) may be treated as a set of ``documents'' for training a text
    embedding model. \textbf{B. Constructing lecture content
    \textit{trajectories}.} After training the model on the sliding windows from
    both lectures, we transform each lecture into a ``trajectory'' through text
    embedding space by joining the embedding coordinates of successive sliding
    windows parsed from its transcript. \textbf{C. Embedding multiple lectures and
    questions in a shared space.} We apply the same model (trained on the two
    lectures' windows) to both lectures, along with the text of each question in
    our pool (Supp. Tab.~\questions), to project them into a shared text embedding space.
    This results in one trajectory per lecture and one coordinate for each
    question. Here, we have projected the 15-dimensional embeddings onto their first
    3 principal components for visualization.}

@@

We hypothesized that a topic model trained on transcripts of the two lectures
should also capture the conceptual knowledge probed by each quiz question. If
indeed the topic model could capture information about the deeper conceptual
content of the lectures (i.e., beyond surface-level details such as particular
word choices), then we should be able to recover a correspondence between each
lecture and questions \textit{about} each lecture. Importantly, such a
correspondence could not solely arise from superficial text matching between
lecture transcripts and questions, since the lectures and questions used
different words. Simply comparing the average topic weights from each lecture
and question set (averaging across time and questions, respectively) reveals a
striking correspondence (Supp. Fig.~\topicWeights). Specifically, the average topic
weights from Lecture~1 are strongly correlated with the average topic weights
from Lecture~1 questions ($r(13) = 0.809,~p < 0.001$, 95\% confidence interval
(CI)~$= [0.633,~0.962]$), and the average topic weights from Lecture~2 are
strongly correlated with the average topic weights from Lecture~2 questions
($r(13) = 0.728,~p = 0.002$, 95\% CI~$= [0.456,~0.920]$). At the same time, the
average topic weights from the two lectures are \textit{negatively} correlated with
their non-matching question sets (Lecture~1 video vs.~Lecture~2 questions: $r(13) =
-0.547,~p = 0.035$, 95\% CI~$= [-0.812, ~-0.231]$; Lecture~2 video vs.~Lecture~1 questions: $r(13)
= -0.612,~p = 0.015$, 95\% CI~$= [-0.874,~-0.281]$), indicating that the topic model
also exhibits some degree of specificity. The full set of pairwise comparisons
between average topic weights for the lectures and question sets is reported in
Supplementary Figure~\topicWeights.

@@

If we are able to accurately estimate a participant's knowledge about the
content tested by a given question, our estimates of their knowledge should carry some
predictive information about whether the participant is likely to answer that
question correctly or incorrectly. We developed a statistical approach to test this claim.
For each question in turn, we used Equation~\ref{eqn:prop} to estimate each
participant's knowledge at the given question's embedding space coordinate,
using all \textit{other} questions that participant answered on the same quiz.
%For each question in turn, for each
%participant, we used Equation~\ref{eqn:prop} to estimate (using all
%\textit{other} questions from the same quiz, from the same participant) the
%participant's knowledge at the held-out question's embedding coordinate.
For
each quiz, we grouped these estimates into two distributions: one for the
estimated knowledge at the coordinates of \textit{correctly} answered
questions, and another for the estimated knowledge at the coordinates of
\textit{incorrectly} answered questions (Fig.~\ref{fig:predictions}). We then
used independent samples $t$-tests to compare the means of these distributions
of estimated knowledge.

@@

We transformed each sliding window's text into a topic vector, and then used
linear interpolation (independently for each topic dimension) to resample the
resulting timeseries to one vector per second. We also used the fitted model to
obtain topic vectors for each question in our pool (see Supp. Tab.~\questions). Taken
together, we obtained a \textit{trajectory} for each video, describing its path
through topic space, and a single coordinate for each question
(Fig.~\ref{fig:sliding-windows}C). Embedding both videos and all of the
questions using a common model enables us to compare the content from different
moments of videos, compare the content across videos, and estimate potential
associations between specific questions and specific moments of video.

@@

Intuitively, $\mathrm{ncorr}(x, y)$ is the correlation between two topic
vectors (e.g., the topic vector from one timepoint in a lecture, $x$, and the
topic vector for one question, $y$), normalized by the minimum and maximum
correlations (across all timepoints $t$ and questions $Q$) to range between 0 and 1,
inclusive. Equation~\ref{eqn:prop} then computes the weighted average
proportion of correctly answered questions about the content presented at
timepoint $t$, where the weights are given by the normalized correlations
between timepoint $t$'s topic vector and the topic vectors for each question.
The normalization step (i.e., using $\mathrm{ncorr}$ instead of the raw
correlations) insures that every question
contributes some non-negative amount to the knowledge estimate.

@@

An important feature of our approach is that, given a trained text embedding
model and participants' quiz performance on each question, we can estimate
their knowledge about \textit{any} content expressible by the embedding
model---not solely the content explicitly probed by the quiz questions, or even
appearing in the lectures. To visualize these estimates
(Fig.~\ref{fig:knowledge-maps}, Supp. Figs.~\individualKnowledgeMapsA,~\individualKnowledgeMapsB,~\individualKnowledgeMapsC,~\individualLearningMapsA,
and~\individualLearningMapsB), we used Uniform Manifold Approximation and
Projection (UMAP; \citealp{McInEtal18a, McInEtal18b}) to construct a 2D projection of the
text embedding space. Sampling the original 100-dimensional space at high
resolution to obtain an adequate set of topic vectors spanning the embedding
space would be computationally intractable. However, sampling a 2D grid is
trivial.
